Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-19 docs/paper.tex
updates to main text

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

We evaluate our model along three sets of parameters: how
closely does our clustering agree with neuroanatomists' knowledge? 
Given two cells, how accurately can our model predict the link
between them? And how closely does the spatial extent (within a layer)
of our identified types agree with the known neuroanatomists.

We find that given the dataset many techniques allow for good
link-predictive accuracy but poor agreement with neuroanatomists. In
fact, when we fit a version of our model fixing the clusterings given
by neuroanatomists \todo{Add something here}. As we incorporate
additional information beyond connectivity into our model (soma
distance, soma depth, synapse density profile) we more closely
approximate human clustering.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). a.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. b.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

@@

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:mos6502}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:mos6502}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

@@

There exist a range of previous approaches to the discovery of
neural microcircuitry \todo{ xxx as many cites as possible}. These
generally involve a great deal of manual labor and ad-hoc
determination of what constitutes a “type” of cell -- to this day
there are disagreements in the literature as to the “true” types in
the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

@@

For the basic link-distance model, we take as input a connectivity
matrix $R$ defining the connections between cell $e_i$ and $e_j$, as
well as a distance function $d(e_i, e_j)$ representing a (physical)
distance between adjacent cells. \todo{Add the cell soma caveat?}  See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter$\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.

>>>>

%\documentclass{nature}
\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

We evaluate our model along three sets of parameters
(Fig.\ref{fig:mouseretina_compare}): how closely does our clustering agree
with neuroanatomists' knowledge?  Given two cells, how accurately can
our model predict the link between them? And how closely does the
spatial extent (within a layer) of our identified types agree with the
known neuroanatomists.

As a first measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B AUC, red).  We find that
given the dataset many techniques allow for good link-predictive
accuracy. All the methods allow decent link prediction with an AUC in
the .9 range. However, our algorithm clearly outperforms the simple
statistical models that only use connectivity.

As a second measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B ARI, blue). We find that our
algorithm far outperforms the controls. We also find that when it is
based on more of the same information used by anatomists use then it
gets better at agreeing with these anatomists. In particular, using
connectivity, distance, synapse distribution and soma depth leads to
the highest ARI. When using the available information the algorithm
produces a good fit to human anatomist judgments.

Finally we look at the spatial extent of the discovered types both
within a layer and between layers (Fig.\ref{fig:mouseretina_compare}
C). We see that, in the absence of distance information, mere
connectivity information results in types which only span a small
region of space -- essentially “local cliques”. Incorporation of
distance information results in types which span the entire extent of
the layer. The depth variance of all models continues to be
substantially larger than that predicted by human anatomists -- future
directions of work include attempting to more strongly encode this
prior belief of laminarity.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:otherdata}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

@@

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:otherdata}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:otherdata}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

@@

There exist a range of previous approaches to the discovery of neural
microcircuitry\autocite{Mountcastle1957, Douglas1991, Bartho2004,
  Freund1998}.  These generally involve a great deal of manual labor
and ad-hoc determination of what constitutes a “type” of cell -- to
this day there are disagreements in the literature as to the “true”
types in the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

@@

For the basic link-distance model, we take as input a connectivity
matrix $R$ defining the connections between cell $e_i$ and $e_j$, as
well as a distance function $d(e_i, e_j)$ representing a (physical)
distance between adjacent cells.  See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter$\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.
